Concurrent Programming in Harmony

### 2021-02-03 book/paper.tex
Fixed typo
Latest

All instructions are atomically executed.
Most instructions pop values from the stack or push values onto the stack.
At first there is one thread, named \texttt{\_\_init\_\_},
which initializes the state.
It starts executing at instruction 0 and keeps executing until
it reaches the last instruction in the program.
In this case, it executes instructions 0 through 5 first.
The last instruction in that sequence is a \texttt{JUMP}
instruction that sets the program counter to 32
(skipping over the code for \texttt{incrementer} method).
The \texttt{\_\_init\_\_} thread then executes instructions
32 through 40 and finishes.
Once initialization completes, any threads that were spawned
(in this case \texttt{incrementer(0)} and \texttt{incrementer(0)})
can run.

@@

\autoref{fig:uplock} showed a faulty attempt at solving mutual
exclusion using a lock.  The problem with the implementation of the
lock is that checking the lock and setting it if it is available is
not \emph{atomic}.  Thus multiple threads contending for the lock
can all ``grab the lock'' at the same time.  While Peterson's
algorithm gets around the problem, it is not efficient, especially
if generalized to multiple threads.
Worse, most modern processors make accessing shared memory locations
non-atomic to allow for efficient caching and other optimizations.
So Peterson's algorithm would not even work on such processors.
Instead, multi-core processors provide
so-called \emph{interlock instructions}:
\index{interlock instruction}%
special machine instructions
that can read memory and then write it in an indivisible step.

@@

\begin{figure}
\begin{code}
\harmonysource{actor}
\end{code}
\caption{\harmonylink{code/actor.hny} A producer/consumer actor.}
\label{fig:actor}
\end{figure}

\begin{figure}
\begin{code}
\harmonysource{actortest}
\end{code}
\caption{\harmonylink{code/actortest.hny} Test code for producer/consumer actor.}
\label{fig:actortest}
\end{figure}

Some programming languages favor a different way of implementing
synchronization using so-called \emph{actors}~\cite{HBS73}.  Actors are
threads that have only private memory and communicate through
\emph{message passing}.
See \autoref{fig:actorpic} for an illustration.
Given that there is no shared memory in the actor model (other than the message
queues, which have built-in synchronization), there is no need
for critical sections.  Instead, some sequential thread owns a particular
piece of data and other threads access it by sending request messages
to the thread and optionally waiting for response messages.  Each thread
handles one message at a time, serializing all access to the data it owns.
As message queues are FIFO (First-In-First-Out), starvation is prevented.

@@

\autoref{fig:actor} shows a Harmony solution to the
``multiple producer / multiple consumer problem'' (with an unbounded buffer)
using an actor (as illustrated in \autoref{fig:actorpic}).
\texttt{pc\_actor} is essentially a matchmaker: it matches
\texttt{.produce} requests with \texttt{.consume} requests.
Sometimes it may have buffered \texttt{.consume} requests and is waiting for
\texttt{.produce} requests, and sometimes it is vice versa.
The variable \textit{balance} specifies what the situation is: if it is positive
then it has buffered \texttt{.produce} requests;
if it is negative
then it has buffered \texttt{.consume} requests.
Note that both \textit{balance} and \textit{requests} are variables that are
local to the \texttt{pc\_actor} thread.

Each request message is a dictionary with two fields.  One of the fields is
\texttt{.type}, which either is \texttt{.produce} or \texttt{.consume}.
In case of \texttt{.produce}, the second field is \texttt{.item} and holds
the data that is produced.
In case of \texttt{.consume}, the second field is \texttt{.queue} and holds
a pointer to the queue where the response message must be enqueued.
\texttt{.produce} messages do not get a response.

The \texttt{pc\_actor} thread has two arguments.  $q$ is a pointer
to the queue on which it receives messages.  \textit{nrequests} is the total
number of requests that it will handle.  We need the latter because in Harmony
all threads are required to terminate.  In a more realistic implementation,
the \texttt{pc\_actor} would be in an infinite loop awaiting requests.

The shared memory consists entirely of message queues.
The \texttt{produce} method takes two arguments: a pointer to the queue
of the \texttt{pc\_actor} and the item.  The \texttt{consume} method
takes two queue pointer arguments.  The first is a pointer to the queue
of the \texttt{pc\_actor}.  The second is a pointer to the queue of the
invoker: it is the queue to which the response must be posted.

\autoref{fig:actortest} shows how this code may be used.  It spawns
\texttt{NITEMS} producer and consumer threads.  Note that in that
case the \texttt{pc\_actor} is expected to receive \texttt{2 * NITEMS}
requests.

\section*{Exercises}
\begin{problems}
\item \label{ex:cltsvr} A popular model is the \emph{client/server} model.
Here a single actor implements some service, like computing the square of a
number.
A client can send a message containing an integer to the server, and the
server returns a response message containing the square of that integer.
Implement this.
\item Actors and message queues are good for building pipelines.
Develop a pipeline that computes Mersenne primes (primes that are one less
than a power of two).  Write four actors:
\begin{enumerate}
\item an actor that generates a sequence of integers 1 through \texttt{N};
\item an actor that receives integers and forwards only those that are prime;
\item an actor that receives integers and forwards only those that are one
less than a power of two;
\item an actor that receives integers but otherwise ignores them.
\end{enumerate}
Configure two versions of the pipeline, one that first checks if a number
is prime and then if it is one less than a power of two, the other
in the opposite order.  Which do you think is better?
\end{problems}

@@

\section*{Exercises}
\begin{problems}
\item \autoref{ex:cltsvr} explored the \emph{client/server model}.  It is popular
in distributed systems as well.
Develop a protocol for a single client and server using the same network
model as for the ABP protocol.
Hint: the response to a request can contain the same sequence number as the
request.
\item Generalize the solution in the previous exercise to multiple clients.
Each client is uniquely identified.  You may either use separate channel pairs
for each client, or solve the problem using a single pair of channels.
\end{problems}

>>>>

All instructions are atomically executed.
Most instructions pop values from the stack or push values onto the stack.
At first there is one thread, named \texttt{\_\_init\_\_},
which initializes the state.
It starts executing at instruction 0 and keeps executing until
it reaches the last instruction in the program.
In this case, it executes instructions 0 through 5 first.
The last instruction in that sequence is a \texttt{JUMP}
instruction that sets the program counter to 32
(skipping over the code for \texttt{incrementer} method).
The \texttt{\_\_init\_\_} thread then executes instructions
32 through 40 and finishes.
Once initialization completes, any threads that were spawned
(in this case \texttt{incrementer(0)} and \texttt{incrementer(1)})
can run.

@@

\autoref{fig:uplock} showed a faulty attempt at solving mutual
exclusion using a lock.  The problem with the implementation of the
lock is that checking the lock and setting it if it is available is
not \emph{atomic}.  Thus multiple threads contending for the lock
can all ``grab the lock'' at the same time.  While Peterson's
algorithm gets around the problem, it is not efficient, especially
if generalized to multiple threads.
Worse, most modern processors make accessing shared memory locations
non-atomic to allow for efficient caching, out-of-order execution,
and other optimizations.
So Peterson's algorithm would not even work on such processors.
Instead, multi-core processors provide
so-called \emph{interlock instructions}:
\index{interlock instruction}%
special machine instructions
that can read memory and then write it in an indivisible step.

@@

\begin{figure}
\begin{code}
\harmonysource{counter}
\end{code}
\caption{\harmonylink{code/counter.hny} An illustration of the actor approach.}
\label{fig:actor}
\end{figure}

Some programming languages favor a different way of implementing
synchronization using so-called \emph{actors}~\cite{HBS73,Agha86}.  Actors are
threads that have only private memory and communicate through
\emph{message passing}.
See \autoref{fig:actorpic} for an illustration.
Given that there is no shared memory in the actor model (other than the message
queues, which have built-in synchronization), there is no need
for critical sections.  Instead, some sequential thread owns a particular
piece of data and other threads access it by sending request messages
to the thread and optionally waiting for response messages.  Each thread
handles one message at a time, serializing all access to the data it owns.
As message queues are FIFO (First-In-First-Out), starvation is prevented.

@@

\autoref{fig:actor} illustrates the actor approach.
There are three client threads that each want to be assigned
a unique identifier from the set $\{ 0, 1, 2 \}$.
Normally one would use a shared 0-initialized counter and a lock.
Each client would acquire the lock, get the value of the counter
and increment it, and release the lock.
Instead, in the actor approach the counter is managed by a
separate server thread.
Each client sends a request to the server, consisting in this case
of simply the queue to which the server must send the response.
The server maintains a local, zero-initialized counter variable.
Upon receiving a request, it returns a response with the value of
the counter and increments the counter.  No lock is required.
The code is tested using the \textit{done} array.

This illustration is an example of the \emph{client/server} model.
Here a single actor implements some service, and clients send request
messages and receive response messages.  The model is particularly
popular in distributed systems, where each actor runs on a separate
machine and the queues are message channels.  For example, the server
can be a web server, and its clients are web browsers.

\section*{Exercises}
\begin{problems}
\item Actors and message queues are good for building pipelines.
Develop a pipeline that computes Mersenne primes (primes that are one less
than a power of two).  Write four actors:
\begin{enumerate}
\item an actor that generates a sequence of integers 1 through \texttt{N};
\item an actor that receives integers and forwards only those that are prime;
\item an actor that receives integers and forwards only those that are one
less than a power of two;
\item an actor that receives integers but otherwise ignores them.
\end{enumerate}
Configure two versions of the pipeline, one that first checks if a number
is prime and then if it is one less than a power of two, the other
in the opposite order.  Which do you think is better?
\end{problems}

@@

\section*{Exercises}
\begin{problems}
\item \autoref{ch:actor} explored the \emph{client/server model}.  It is popular
in distributed systems as well.
Develop a protocol for a single client and server using the same network
model as for the ABP protocol.
Hint: the response to a request can contain the same sequence number as the
request.
\item Generalize the solution in the previous exercise to multiple clients.
Each client is uniquely identified.  You may either use separate channel pairs
for each client, or solve the problem using a single pair of channels.
\end{problems}
